---
title: "HAI Analysis v5: QC and Univariate Analyses"
author: "Matthew Chung"
date: "9/22/2021"
output: 
  html_document:
    toc: true
    theme: united
---

# Preanalysis

## Load packages and view sessionInfo

```{r,warning=FALSE,message = FALSE}
library(biomaRt)
library(cowplot)
library(clusterProfiler)
library(cowplot)
library(DESeq2)
library(edgeR)
library(enrichplot)
library(ggplot2)
library(gprofiler2)
library(openxlsx)
library(pathview)
library(reshape2)
library(see)
library(splines)
library(SuperExactTest)
library(UpSetR)
library(readr)
library(dplyr)
library(forcats)
```

#MRH 26May2022: 
##Upload the .RData file made in preprocessing to get data that's already undergone necessary processing.

```{r}
load("../data.RData")
tpm.hemoglobin <- read.csv("../tpm.hemoglobin.csv", row.names = 1)



metadata$log2sero <- log2(metadata$average_seroconversion)
metadata$seroconversion_2cat <- as.factor(ifelse(metadata$log2sero >= 2,"high","low"))
metadata$D0_score_2cat <- as.factor(ifelse(metadata$Composite_Baseline >= 25, "high", "low"))
```

## Set metadata covariates
###MRH Edit: Removed "Composite_SeroConversion", "CorrectedSeroconversion_composite" and added "log2sero", 


```{r}
covar1 <- c("log2sero","Run.ID","Age","Gender","BMI_Value", 
            "Race_Ethnicity","Month.vaccinated","Composite_Baseline","Composite_D28","Prevacc.naive")
```

# Data QC: Distribution of gene counts per sample

## Plot gene count distribution

```{r,fig.height=4,fig.width=2}
figs4 <- ggplot()+
  geom_violinhalf(aes(x="A",y=colSums(counts.subset)),fill="#f8766d")+
  geom_boxplot(aes(x="A",y=colSums(counts.subset)),fill="#f8766d",width=0.2)+
  geom_jitter(aes(x="A",y=colSums(counts.subset)),alpha=0.5)+
  labs(y="protein-coding gene counts")+
  theme_bw()+
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

png(filename = "figs6.png",
    height = 2000,
    width = 800,
    units = "px",
    res = 300)
figs4
dev.off()

pdf(paste0("figs6.pdf"),
    height = 25,
    width = 10)
figs4
dev.off()
```

# MDS/PC plot showing major sources of variation (Batch effect, sex)

## Set plotting functions

```{r}
plot_mds <- function(mds, comp1, comp2, groups){
  plot.df <- as.data.frame(cbind(mds$points[,comp1],
                                 mds$points[,comp2]))
  # imp1 <- summary(pca)$importance[2,comp1]
  # imp2 <- summary(pca)$importance[2,comp2]
  
  set.seed(2)
  plot.df <- plot.df[sample(1:nrow(plot.df)),]
  set.seed(2)
  groups <- groups[sample(1:nrow(plot.df))]
  
  
  plot <- ggplot()+
    geom_point(aes(x=!!plot.df[,1],y=!!plot.df[,2],color=!!groups))+
    labs(x=paste0("MDS",comp1),
         y=paste0("MDS",comp2))+
    guides(color = "none")+
    theme_bw()
  
  if(type(groups) == "double"){
    plot <- plot+
      scale_color_gradient2(mid = "blue",high="orange")
  }
  
  return(plot)
}

plot_pca <- function(pca, comp1, comp2, groups){
  plot.df <- as.data.frame(cbind(pca$x[,comp1],
                                 pca$x[,comp2]))
  imp1 <- summary(pca)$importance[2,comp1]
  imp2 <- summary(pca)$importance[2,comp2]
  
  set.seed(2)
  plot.df <- plot.df[sample(1:nrow(plot.df)),]
  set.seed(2)
  groups <- groups[sample(1:nrow(plot.df))]
  
  
  plot <- ggplot()+
    geom_point(aes(x=!!plot.df[,1],y=!!plot.df[,2],color=!!groups))+
    labs(x=paste0("PC",comp1," (",round(imp1*100,2),"%)"),
         y=paste0("PC",comp2," (",round(imp2*100,2),"%)"))+
    guides(color = "none")+
    theme_bw()
  
  if(type(groups) == "double"){
    plot <- plot+
      scale_color_gradient2(mid = "blue",high="orange")
  }
  return(plot)
}

plot_legend <- function(name,groups){
  plot <- ggplot()+
    geom_point(aes(x="1",y="1",color=!!groups))+
    labs(color=name)+
    theme_bw()
  
  if(type(groups) == "double"){
    plot <- plot+
      scale_color_gradient2(mid = "blue",high="orange")
  }
  legend <- get_legend(plot)
  return(legend)
}
```

## Plot legends
```{r,fig.width=12,fig.height=6}
plot.list <- list()
for(i in 1:length(covar1)){
  #legend <- ggdraw(plot_legend(covar1[i],groups=metadata[,colnames(metadata) == covar1[i]]))
  
  plot.list[[i]] <- plot_legend(covar1[i],groups=metadata[,colnames(metadata) == covar1[i]])
  
}
plot_grid(plotlist = plot.list,ncol=5,align='hv')
```

## Run and plot PCA

```{r,fig.height=20,fig.width=12}
covartitles <- c("Log2 Seroconversion", "Run ID", "Age", "Gender", "BMI", "Race/Ethnicity", "Month Vaccinated", "Average Baseline HAI", "Average D28 HAI", "Prevaccination Status")

pca.df <- tpm.subset
# pca.df <- t(scale(t(log2(tpm.subset+1))))

pca <- prcomp(t(pca.df),scale. = TRUE)
#pca <- prcomp(t(pca.df), center = TRUE,scale. = TRUE)

plot.list <- list()
for(i in 1:length(covar1)){
  plot.list <- append(plot.list,
                     list(plot_pca(pca,1,2,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covartitles[i]),
                          plot_pca(pca,2,3,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covartitles[i]),
                          plot_pca(pca,3,4,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covartitles[i]),
                          plot_pca(pca,4,5,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covartitles[i]),
                          plot_pca(pca,5,6,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covartitles[i])))
}

plot_grid(plotlist = plot.list,ncol=5,align='hv')

figs8 <- plot_grid(plotlist = plot.list,ncol=5,align='hv')

png(filename = "figs8.png",
    height = 5500,
    width = 4000,
    units = "px",
    res = 300)
figs8
dev.off()

pdf(paste0("figs8.pdf"),
    height = 17,
    width = 12)
figs8
dev.off()

```

## Run and plot MDS

```{r,fig.height=20,fig.width=12}
mds <- cmdscale(dist(t(pca.df)), k = 10, eig=TRUE)

plot.list <- list()
for(i in 1:length(covar1)){
  plot.list <- append(plot.list,
                     list(plot_mds(mds,1,2,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covar1[i]),
                          plot_mds(mds,2,3,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covar1[i]),
                          plot_mds(mds,3,4,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covar1[i]),
                          plot_mds(mds,4,5,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covar1[i]),
                          plot_mds(mds,5,6,groups=metadata[,colnames(metadata) == covar1[i]])+
                             labs(title=covar1[i])))
}

plot_grid(plotlist = plot.list,ncol=5,align='hv')
```

# Cumulative probability plot showing fraction of total reads (y-axis) vs ranked genes (x axis)

## Plot cumulative probability plot

```{r}
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))


top_exp_genes <- rownames(tpm.hemoglobin)[rev(order(rowMeans(tpm.hemoglobin)))][1:20]

plot.df <- tpm.hemoglobin[rev(match(top_exp_genes,rownames(tpm.hemoglobin))),]
plot.df <- melt(as.matrix(plot.df))

genenames <- as.data.frame(matrix(nrow = nrow(plot.df),
                                 ncol = 2))
geneids <- as.character(unique(plot.df$Var1))
genenames <- as.data.frame(getBM(filters= "ensembl_gene_id", 
                  attributes = c("ensembl_gene_id", "hgnc_symbol"),
                  values=geneids,
                  mart= mart))

if(genenames$hgnc_symbol %in% ""){genenames$hgnc_symbol <- "NA"}


plot.df$names <- as.factor(genenames$hgnc_symbol[match((plot.df$Var1), genenames$ensembl_gene_id)])
for(i in 1:nrow(plot.df)){ 
  plot.df$names <- as.character(plot.df$names)
  plot.df$Var1 <- as.character(plot.df$Var1)
  if(plot.df[i,4] ==""){plot.df[i,4] <- plot.df[i,1]}
}

plot.df$names <- as.factor(plot.df$names)
plot.df$Var1 <- as.factor(plot.df$Var1)

high_exp_hemo_genes <- c("ENSG00000188536","ENSG00000244734","ENSG00000206172")
hemo_genes <- c("HBA2", "HBB", "HBA1")

boxfill <- ifelse(levels(plot.df[,1]) %in% high_exp_hemo_genes,"red",NA)


summ <- plot.df %>% group_by(names) %>% summarize(mean = mean(value))

plot.df$summ <- summ$mean[match(plot.df$names, summ$names)]

plot.df <-  arrange(plot.df, plot.df$summ) 
boxfill_hemo <- ifelse(levels(plot.df$names) %in% hemo_genes,"red",NA)


figs7 <- ggplot()+
  geom_boxplot(aes(x=plot.df[,3],
                   y=plot.df[,4]),
                   fill = boxfill_hemo)+
  labs(x="average TPM",y="")+
  scale_x_continuous(trans="log10")+
  theme_bw()

figs7

png(filename = "figs7.png",
    height = 2000,
    width = 2000,
    units = "px",
    res = 300)
figs7
dev.off()

pdf(paste0("figs7.pdf"),
    height = 10,
    width = 10)
figs7
dev.off()
```

