---
title: "HAI Analysis Fig 6"
author: "Matthew Chung"
date: "10/16/2022"
output: 
  html_document:
    toc: true
    theme: united
---
# Preanalysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(extrafont)
library(remotes)
remotes::install_version("Rttf2pt1", version = "1.3.8")
extrafont::font_import()
```

```{r}
library(biomaRt)
library(cowplot)
library(clusterProfiler)
# library(DESeq2)
library(edgeR)
library(enrichplot)
#library(gameofthrones)
library(ggpubr)
#library(GOSemSim)
library(ggplot2)
library(ggridges)
# library(gprofiler2)
# library(openxlsx)
# library(pathview)
library(reshape2)
# library(see)
library(splines)
# library(SuperExactTest)
library(UpSetR)
library(png)
library(ggpubr)
library(dplyr)
library(stringr) 
library(gridExtra)
library(egg)
library(patchwork)
#library(extrafont)
```


##Load the Data and log2 seroconversion

```{r}
load("C:/Users/chungm6/Desktop/projects/david_gresham_01/data.RData")

metadata$log2sero <- log2(metadata$average_seroconversion)

metadata$log2baseline <- log2(metadata$Composite_Baseline)

```

## Add metadata readout for categorizing average_seroconversion and day 0 serostatus
```{r}
metadata$seroconversion_2cat <- as.factor(ifelse(metadata$log2sero >= 2,"high","low"))
metadata$D0_score_2cat <- as.factor(ifelse(metadata$Composite_Baseline >= 25, "high", "low"))

sum(grepl("low", metadata$seroconversion_2cat))
```
# DE analysis

##Set covariates of interest
These include covariates that are well distributed across the subjects:
1) BMI
2) Gender
3) Age
4) Average seroconversion
5) Baseline HAI
```{r}
covariates <- c("BMI_Value", "Gender", "Age", "log2sero", "log2baseline")

covariates_titles <- c("BMI", "Sex", "Age", "Seroconversion", "Baseline HAI")

#metadata$Seroconversion_bivar <- ifelse(metadata$Composite_SeroConversion > 8, "high", "low")
```


## Calculate Differential Expression based on selected covariates
```{r}
FDR.cutoff <- 0.05
degenes.edgeR <- list()
toptags <- list()

for(j in 1:length(covariates)){
  group <- metadata[,which(colnames(metadata) == covariates[j])]
  
  if(length(unique(group)) > 10){
    spline <- ns(group, df = 1)
    design <- model.matrix(~spline)
  }else{
    design <- model.matrix(formula(paste("~",covariates[j])),data = metadata)
  }
  
  y <- DGEList(counts = counts.subset,
               group = group)
  
  keep <- filterByExpr(y)
  
  y <- y[keep, ,keep.lib.sizes = F]
  y <- calcNormFactors(y, method = "TMM")
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  
  qlf <- glmQLFTest(fit, coef = 2:ncol(design))
  degenes <- topTags(qlf, n = Inf)
  toptags[[j]] <- degenes
  degenes.edgeR[[j]] <- rownames(degenes$table)[degenes@.Data[[1]]$FDR < FDR.cutoff]
  
}

names(degenes.edgeR) <- covariates_titles

```

## Plot UpSet plot of Differentially expressed genes

```{r}
fig2 <- list()
fig2$a<- upset(fromList(degenes.edgeR), 
                order.by = "freq", 
                nsets = length(covariates),
                mainbar.y.max = 3000, 
                text.scale = 5, 
                mainbar.y.label = "Intersection Size", 
                sets.x.label = "Differentially\nExpressed Genes", 
                point.size = 6,
                line.size = 1, 
                nintersects = 12,
               set_size.show = TRUE)


x1 <- unlist(degenes.edgeR, use.names = FALSE)
rownames(fig2$a$New_data) <- x1[ !duplicated(x1) ]

sero_exclusive <- rownames(fig2$a$New_data)[fig2$a$New_data$Seroconversion == 1 & rowSums(fig2$a$New_data) == 1]
```


## Switch TPM rownames to gene names and subset by log2sero DE genes

```{r}
ensembl <- useEnsembl(biomart = "genes")
ensembl <- useDataset(dataset = "hsapiens_gene_ensembl", mart = ensembl)
biomart <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name'),
                filters = 'ensembl_gene_id',
                values = degenes.edgeR$Seroconversion, 
                mart = ensembl)
biomart <- biomart[match(degenes.edgeR$Seroconversion,biomart[,1]),]
biomart[,1] <- degenes.edgeR$Seroconversion

biomart[,2] <- ifelse(biomart[,2] == "" | biomart[,2] == "Metazoa_SRP",biomart[,1],biomart[,2])

biomart[,2] <- make.names(biomart[,2],unique = T)

tpm.subset <- tpm.subset[degenes.edgeR$Seroconversion,]

rownames(tpm.subset) <- biomart[,2]
degenes.edgeR$Seroconversion <- biomart[,2]
sero_exclusive <- biomart[match(sero_exclusive,biomart[,1]),2]
```


## Load modules

```{r}
modules <- list(PC1=read.delim("tpm_hemoglobin_edgeR_log2CPM_RundIDadj_anno_0q5__FDR0.05_modules_PC1.tsv"),
                PC2=read.delim("tpm_hemoglobin_edgeR_log2CPM_RundIDadj_anno_0q5__FDR0.05_modules_PC2.tsv"),
                PC3=read.delim("tpm_hemoglobin_edgeR_log2CPM_RundIDadj_anno_0q5__FDR0.05_modules_PC3.tsv"),
                PC4=read.delim("tpm_hemoglobin_edgeR_log2CPM_RundIDadj_anno_0q5__FDR0.05_modules_PC4.tsv"),
                PC5=read.delim("tpm_hemoglobin_edgeR_log2CPM_RundIDadj_anno_0q5__FDR0.05_modules_PC5.tsv"))

top_modules <- PC1$Module[order(PC1$log2Sero.rnk)][1:20]

for(i in 1:length(modules)){
  rownames(modules[[i]]) <- modules[[i]]$Module
  modules[[i]] <- modules[[i]][top_modules,metadata$File.Name]
}
```


## Create random forest model

### Split data into training and testing sets

```{r}
covar2 <- c("Age","Gender","BMI_Value","Prevacc.naive","D0_score_2cat")

seeds <- c(80:89)
rf.alldata <- list(PD=as.data.frame(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2]),
                   PD_DEGs=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                t(tpm.subset[sero_exclusive,]))),
                   PD_DEGs_PC1=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[sero_exclusive,]),
                                                    t(modules$PC1))),
                   PD_DEGs_PC2=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[sero_exclusive,]),
                                                    t(modules$PC2))),
                   PD_DEGs_PC3=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[sero_exclusive,]),
                                                    t(modules$PC3))),
                   PD_DEGs_PC4=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[sero_exclusive,]),
                                                    t(modules$PC4))),
                   PD_DEGs_PC5=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[sero_exclusive,]),
                                                    t(modules$PC5))))

rf.traindata <- list()
rf.testdata <- list()

library(caTools)
for(i in 1:length(rf.alldata)){
  rf.traindata[[i]] <- list()
  rf.testdata[[i]] <- list()
  for(j in 1:length(seeds)){
    set.seed(seeds[j])
    sample = sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    rf.traindata[[i]][[j]] = rf.alldata[[i]][sample == T,]
    rf.testdata[[i]][[j]]  = rf.alldata[[i]][sample == F,]
  }
}
names(rf.traindata) <- names(rf.alldata)
names(rf.testdata) <- names(rf.alldata)
```


### Tune random forest model for optimal number of variables sampled at each split

```{r,warning=FALSE,message = FALSE}
rf.mdl <- list()

library(randomForest)
for(i in 1:length(rf.alldata)){
  rf.mdl[[i]] <- list()
  for(j in 1:length(seeds)){
    set.seed(seeds[j])
    sample <- sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    set.seed(seeds[j])
    mtry <- tuneRF(x = rf.traindata[[i]][[j]],
                   y = metadata$seroconversion_2cat[sample == T],
                   mtryStart=round(sqrt(ncol(rf.traindata[[i]][[j]]))),
                   ntreeTry=500,
                   stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
    best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
  
    set.seed(seeds[j])
    sample = sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    set.seed(seeds[j])
    rf.mdl[[i]][[j]] <-randomForest(x = rf.traindata[[i]][[j]],
                                y = metadata$seroconversion_2cat[sample == T],
                                mtry=best.m,
                                ntree=500,
                                proximity =T,
                                importance = T)
  }
}
```

### Plot average OOB error

```{r,fig.height=3,fig.width=6}
oob.error <- as.data.frame(matrix(nrow=0,
                                  ncol=2))
colnames(oob.error) <- c("dataset","errrate")
for(i in 1:length(rf.mdl)){
  oob.error <- as.data.frame(rbind(oob.error,
                                   cbind(names(rf.alldata)[i],
                                         unlist(lapply(rf.mdl[[i]],function(x){x$err.rate[500,1]})))))
  
  
}

plot.df <- oob.error

plot.df[,2] <- as.numeric(as.character(plot.df[,2]))

plot.df[,1] <- factor(plot.df[,1],
                      levels=rev(unique(plot.df[,1])))

oob.plot <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[,2],
                     y=!!plot.df[,1],
                     fill=!!plot.df[,1]),
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[,2],
                  y=!!plot.df[,1]))+
    guides(fill = 'none')+
    labs(x="OOB estimate of error rate",y="")+
    theme_bw()

oob.plot
```


### Plot boxplot of AUC values from ROC curves

```{r,fig.height=3,fig.width=6}
auc <- c()
for(i in 1:length(rf.mdl)){
  for(j in 1:length(rf.mdl[[i]])){
    set.seed(seeds[j])
    sample <- sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    # rocobj <- roc(metadata$seroconversion_2cat[sample == F],
    #               predict(rf.mdl[[i]][[j]], newdata=rf.testdata[[i]][[j]],type="prob")[,1],
    #               smooth=T)
    auc <- c(auc,
             auc(metadata$seroconversion_2cat[sample == F],
                  predict(rf.mdl[[i]][[j]], newdata=rf.testdata[[i]][[j]],type="prob")[,1]))
  }
}
plot.df <- as.data.frame(cbind(oob.error[,1],
                               auc))

plot.df[,2] <- as.numeric(as.character(plot.df[,2]))
plot.df[,1] <- factor(plot.df[,1],
                      levels=rev(unique(plot.df[,1])))

auc.boxplot <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[,2],
                     y=!!plot.df[,1],
                     fill=!!plot.df[,1]),
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[,2],
                  y=!!plot.df[,1]))+
  guides(fill='none')+
    labs(x="AUC",y="")+
    theme_bw()

auc.boxplot

aggregate(.~V1,data=plot.df[,1:2],mean)

```

### Run crossvalidation for random forest

```{r,fig.height=5,fig.width=11}
library(rfUtilities)
colors <- c("#FB61D7","#A58AFF","#00B6EB","#00C094","#53B400","#C49A00","#F8766D")

plots  <- list()
for(i in 1:length(rf.mdl)){
  set.seed(86)
  rf.cv <- rfcv(trainx = rf.alldata[[i]], 
                trainy = metadata$seroconversion_2cat, 
                mtryStart=round(sqrt(ncol(rf.traindata[[i]][[1]]))),
                cv.fold=10)
  plot <- ggplot(mapping=aes(x=!!rf.cv$n.var,
                                         y=!!rf.cv$error.cv))+
                            geom_point(color=colors[i],size=2)+
                            geom_line(color=colors[i],size=1)+
                            labs(title=names(rf.alldata)[i],x="number of variables",y="cross validation error")+
                            scale_x_continuous(trans="log10")+
                            theme_bw()
  plots[[i]] <- plot
  print(i)
}
names(plots) <- names(rf.alldata)
xval.plot <- plots

```

```{r,fig.height=8,fig.width=11}
s16A <- egg::ggarrange(oob.plot,
                      auc.boxplot+theme(axis.text.y=element_blank()),
               ncol=2)
s16B <- egg::ggarrange(plots = xval.plot,
               ncol=3)

pdf(paste0("fig_s17.pdf"),
    height = 8,
    width = 11)

plot_grid(s16A,s16B,
          rel_heights = c(1,3),
          labels = c("A","B"),
          ncol = 1)
dev.off()

```

### Plot random forest variable importance

```{r,fig.height=15,fig.width=9}
colors <- c("#FB61D7","#A58AFF","#00B6EB","#00C094","#53B400","#C49A00","#F8766D")

acc_plots.mda.list <- list()
acc_plots.mdg.list <- list()

for(i in 1:length(rf.mdl)){
  importance.df <- do.call(cbind,lapply(rf.mdl[[i]],function(x){return(x$importance)}))
  rownames(importance.df) <- gsub("Gender","Sex",rownames(importance.df))
  rownames(importance.df) <- gsub("BMI_Value","BMI",rownames(importance.df))
  rownames(importance.df) <- gsub("Prevacc.naive","Previous Vaccination",rownames(importance.df))
  rownames(importance.df) <- gsub("D0_score_2cat", "Baseline HAI", rownames(importance.df))
  
  acc.order <- rownames(importance.df)[order(-rowMeans(importance.df[,colnames(importance.df) == "MeanDecreaseAccuracy"]))]
  gini.order <- rownames(importance.df)[order(-rowMeans(importance.df[,colnames(importance.df) == "MeanDecreaseGini"]))]
  
  if(i!=1){
    acc.order <- acc.order[1:10]
    gini.order <- gini.order[1:10]
  }
  
  plot.df <- melt(importance.df)
  plot.df <- plot.df[plot.df[,1] %in% acc.order,]
  plot.df[,1] <- factor(plot.df[,1],levels=rev(acc.order))
  
  acc_plots.mda.list[[i]] <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",3],
                     y=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",1]),
                 fill=colors[i],
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",3],
                   y=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",1]))+
    labs(x="Mean Decrease Accuracy",y="",title = names(rf.alldata)[i])+
    theme_bw()
  
  plot.df <- melt(importance.df)
  plot.df <- plot.df[plot.df[,1] %in% gini.order,]
  plot.df[,1] <- factor(plot.df[,1],levels=rev(gini.order))

  acc_plots.mdg.list[[i]] <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseGini",3],
                     y=!!plot.df[plot.df[,2] == "MeanDecreaseGini",1]),
                 fill=colors[i],
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseGini",3],
                   y=!!plot.df[plot.df[,2] == "MeanDecreaseGini",1]))+
    labs(x="Mean Decrease Gini",y="",title = names(rf.alldata)[i])+
    theme_bw()
}
names(acc_plots.mda.list) <- names(rf.alldata)
names(acc_plots.mdg.list) <- names(rf.alldata)

s17a <- egg::ggarrange(plots = acc_plots.mda.list,ncol=1)
s17b <- egg::ggarrange(plots = acc_plots.mdg.list,ncol=1)
s17 <- plot_grid(s17a, 
                 s17b,
                    ncol=2)

pdf(paste0("fig_s18.pdf"),
    height = 15,
    width = 9)
s17
dev.off()
```