---
title: "Figures 4, S14, S15, S16"
author: 
  - name: "Christian Forst"
    email: chris@santafe.edu
date: "6/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MEGENA Analysis Pipeline

## Load packages and view sessionInfo

```{r,warning=FALSE,message = FALSE}
library(biomaRt)
library(ggplot2)
library(gprofiler2)
library(reshape2)
library(SuperExactTest)
library(EnhancedVolcano)
```

## Initiating additional fuctions

```{r Initialization}
options(stringsAsFactors=F)
```

## Load data

```{r}
load("Fig_4.RData")
```



## Correlation preprocessing

```{r}
covar <- c("average_log2Baseline", "average_log2D28", "average_log2SC")

metadata <- metaNew

f <- rnaseq.all.logCPM
rownames(f) <- f[,1]

rownames(metadata) <- metadata$File.Name
trait0 <- metadata

common.samples = intersect(colnames(f), rownames(trait0))

geno.xT <- t(f[,common.samples])
bg  <- colnames(geno.xT)
bg.g<- unique(genes[bg,"hgnc_symbol"])

trait <- trait0[common.samples,covar]

```

## Run correlation analysis

```{r Correlation}
indir  <- "MEGENA/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05/"
outdir <- indir
tmfile <- paste0(indir, "multiscale_significant.modules.txt") # the default MEGENA module file
tmbase <- paste0(outdir, rev(strsplit(indir, split="/")[[1]])[1])         # name your output file

msm <- readSigModules(tmfile)
```
## Do some correlation analysis

### First just calculate the 1st principal component 

```{r PC, echo=F, message=FALSE, warning=FALSE}
PCs <- NULL
for (i in 1:length(msm)) {
  message("  Processing module ", names(msm)[i])
  gs <- intersect(colnames(geno.xT), msm[[i]])
  x.modT <- geno.xT[, gs]   # select the corresponding genes from module
  x.pca  <-prcomp(x.modT)$x[,1]
  x.pca.sign<-sign(sum(cor(x.pca, x.modT, use="na.or.complete")))
  if ( is.na(x.pca.sign)) { next }
  x.pca <- x.pca * x.pca.sign
  tmp <- data.frame(Module=gsub("c1_", "M", names(msm)[i]), module=names(msm)[i], t(x.pca))
  PCs <- rbind(PCs, tmp)
}
write.table(PCs, file=paste0(tmbase,"_modules_PC1.tsv"), row.names=F, quote=F, sep="\t")
```

### Second, correlate 1st PC with trait (for now ignoring the calculatin of the 1st PC from above)

```{r Corr2, warning=FALSE, echo=F, message=F}
 message("Calculate MEGENA correlation")
#MEGENA
 trait.pcaSc <- NULL
 trait.pcaSp <- NULL
 for (i in 1:length(msm)) {
  message("  Processing module ", names(msm)[i])
  gs <- intersect(colnames(geno.xT), msm[[i]])
  x.modT <- geno.xT[, gs]   # select the corresponding genes from module
  x.pca  <-prcomp(x.modT)$x[,1]
  x.pca.sign<-sign(sum(cor(x.pca, x.modT, use="na.or.complete")))
  if ( is.na(x.pca.sign)) { next }
  x.pca <- x.pca * x.pca.sign  
  x.cor  <- apply(trait, 2, function(x) cor.test(x.pca,x,use="na.or.complete",method="spearman")$estimate)
  x.corp <- apply(trait, 2, function(x) cor.test(x.pca,x,use="na.or.complete",method="spearman")$p.value)
  trait.pcaSp <- rbind(trait.pcaSp, c(names(msm)[i],x.corp))  
  trait.pcaSc <- rbind(trait.pcaSc,  c(names(msm)[i],x.cor))
}
q <- apply(trait.pcaSp[,-1], 2, function(x) p.adjust(as.numeric(x), method="fdr")) 
trait.pcaSq <- data.frame(trait.pcaSc[,1], q)
colnames(trait.pcaSq) <- colnames(trait.pcaSp)
```

## Plot correlations using EnhancedVolcanoPlot

```{r,fig.height=5,fig.width=8}
ePlot <- list()

for (i in 2:ncol(trait.pcaSc)) {
  tmp <- data.frame(module=trait.pcaSc[,1], Module=gsub("c1_", "M", trait.pcaSc[,1]), corr=as.numeric(trait.pcaSc[,i]), pval=as.numeric(trait.pcaSp[,i]), qval=as.numeric(trait.pcaSq[,i]))
  eP <- EnhancedVolcano(tmp, tmp$Module, "corr", "pval", pCutoff = 0.05, FCcutoff = 0.1, labSize = 5, xlab="Spearman Corr.", drawConnectors = T, legendLabels=c('NS', 'Corr.', 'P', 'P & Corr.'), title=colnames(trait.pcaSc)[i], xlim=c(-.65,.65))
  ePlot <- c(ePlot, list(eP))
}
plot_grid(plotlist=ePlot[1:3], ncol=3, nrow=1)
#plot_grid(plotlist=ePlot[4:6], ncol=3, nrow=1)
```


## Do the same with long correlation version

```{r,fig.height=5,fig.width=12}
ePlot <- list()

#trait.long <- trait.pcaS.long
trait.long <- tt2

trait.long[grep("Base",trait.long$Trait),"Trait"] <- "Baseline HAI"
trait.long[grep("D28",trait.long$Trait),"Trait"]  <- "Response HAI"
trait.long[grep("SC",trait.long$Trait),"Trait"]   <- "Seroconversion"

ts <- unique(trait.long$Trait)

for (i in 1:3) {
  tmp <- trait.long[which(trait.long$Trait==ts[i]),]
  eP <- EnhancedVolcano(tmp, tmp$V1, "corr", "qval", pCutoff = 0.05, FCcutoff = 0.1,  labSize = 3, xlab="Spearman Corr.", drawConnectors = T, legendLabels=c('NS', 'Corr.', 'P', 'P & Corr.'), legendLabSize=10, title=ts[i], xlim=c(-.65,.65))
  ePlot <- c(ePlot, list(eP))
}
ePG1 <- plot_grid(plotlist=ePlot[1:2], ncol=2, nrow=1)
plot_grid(plotlist=ePlot[1:3], ncol=3, nrow=1)
ggsave("Papers/Figures/modules_avLog2Sero_qval_EnhV2.png")
#plot_grid(plotlist=ePlot[4:6], ncol=3, nrow=1)
```

## Write separate correlation tables

```{r}
tname ="avLog2Sero"

write.table(trait.pcaSc, file=paste0(tmbase, "_", tname, "_pcaSc.tsv"), row.names=F, quote=F, sep="\t")
write.table(trait.pcaSp, file=paste0(tmbase, "_", tname, "_pcaSp.tsv"), row.names=F, quote=F, sep="\t")
write.table(trait.pcaSq, file=paste0(tmbase, "_", tname, "_pcaSq.tsv"), row.names=F, quote=F, sep="\t")

```

## Reshape and processing for bubble plots

```{r}
# reshape
tmp <-data.frame(trait.pcaSc)
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
traitc.long <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
#
tmp <-data.frame(trait.pcaSp)
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
colnames(tmp)[1] <- "V1"
traitp.long <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
#
tmp <-data.frame(trait.pcaSq)
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
colnames(tmp)[1] <- "V1"
traitq.long <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
traits <- cbind(traitc.long, pval=traitp.long[,3], qval=traitq.long[,3])
traits$V1 <- gsub("c1_", "M", traits$V1)

write.table(traits, file=paste0(tmbase, "_", tname, "_long.tsv"), row.names=F, quote=F, sep="\t")

```

## Doing bubble plots

```{r,fig.height=6,fig.width=8}
tname <- "Matt_all_avSero_pval"
#ix <- which(traits$qval <=0.27); length(ix)
ix <- which(abs(traits$corr) >=0.13); length(ix)
t2 <- traits[ix,]
ggplot(data.frame(t2), aes(x=Trait, y=factor(V1))) + geom_point(aes(color=corr, size= -log10(qval))) +
    scale_color_distiller(palette="RdYlBu", direction=-1, limits=c(-.41,.41)) + theme_bw() + labs(x="Trait", y="Modules", title=tname) +
    theme_bw() + theme(text = element_text(size = 16), axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
#ggsave(paste0(outdir,"module_", tname, "_corr_bubblePlot.png"), width = 8, height = 7)
#ggsave(paste0(outdir,"module_", tname, "_corr_bubblePlot.pdf"), width = 8, height = 7)

```

## Prepare for module enrichment

```{r}
DEGs <- avSero.list
#DEGs <- DEGs.PConly
y <- msm
b <- lapply(y, function(x) unique(genes[x,"hgnc_symbol"]))
b <- lapply(b, setdiff, y="")

tID <- "DEMatt_avLog2SeroRank"
tID2<- "DEMSigDBenrich_avLog2SeroRank"

```

## Perform MEGENA enrichment

```{r}
tm.DEG <- NULL
for (d in DEGs) {    # use DEGs list
  o <- coexEnrich(y, d, bg)
  tm.DEG <- cbind(tm.DEG, o)
}
cp <- apply(tm.DEG[,grep("p.value", colnames(tm.DEG))], 2, as.numeric)
#cp <- apply(tm.DEG[,grep("corr.p.val", colnames(tm.DEG))], 2, as.numeric)
ix<-apply(cp, 2, function(x) all(is.na(x)))
cp <- cp[,which(!ix)]  # remove all NA columns
if ( is.null(dim(cp))) {next}   # one-dim cp
colnames(cp) <- names(DEGs)[!ix]
rnk <- rankOrder(cp)
module <- sub("c1_", "M", tm.DEG[,1])
tmp <- cbind(tm.DEG[,1], module, rnk, cp) # cannot include cp matrix as # of DEGs differ
#  tm.cp.rank <- rbind(tm.cp.rank, tmp)
fn2 <- paste0(tmbase, "_", tID, ".tsv")
write.table(tmp, file=fn2, row.names=F, quote=F, sep="\t")
```


## Do some more ranking using combined serological info

```{r}
cp2 <- apply(cbind(cp[,grep("Composite", colnames(cp))], trait.pcaSp[,grep("Composite", colnames(trait.pcaSp))]),2 , as.numeric)
r2 <- rankOrder(cp2)
```


## Assemble matrix

```{r}
# use 'out' from the importedmodule enrichment table (above)
t2 <- trait.pcaSc
colnames(t2) <- paste0(colnames(t2), ".corr")
out <- cbind(out, t2)
t2 <- trait.pcaSp
colnames(t2) <- paste0(colnames(t2), ".pval")
out <- cbind(out, t2)
t2 <- trait.pcaSq
colnames(t2) <- paste0(colnames(t2), ".qval")
out <- cbind(out, t2)
#out <- data.frame(Sero.rnk=r2,out)
out <- data.frame(avSero.rnk=rnk,out)

fn3 <- paste0(tmbase, "_", tID2, "_traitCorr.tsv")
write.table(out, file=fn3, row.names=F, quote=F, sep="\t")

```

## make some plots
### Do some reshaping first

```{r}
num <- 10  # # to display
# Careful! some out columns are not numeric!
o <- order(out$avSero.rnk); ptt.t <- ""
#o <- order(as.numeric(out$Age.pval));  ptt.t <- ""
#o <- order(abs(out$Age.corr));  ptt.t <- ""
#o <- order(as.numeric(out$BMI.pval));  ptt.t <- ""
#o <- order(out$Composite_SeroConversion.pval)
#o <- order(out$Composite_D28.pval)
m <- out[o[1:num],"module"]

# reshape
tmp <-data.frame(trait.pcaSc)
rownames(tmp) <- gsub("c1_", "M", tmp[,1])
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
#tmp <- tmp[o[1:num],]
tmp <- tmp[m,]
ttc <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
#
tmp <-data.frame(trait.pcaSp)
rownames(tmp) <- gsub("c1_", "M", tmp[,1])
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
#tmp <- tmp[o[1:num],]
tmp <- tmp[m,]
colnames(tmp)[1] <- "V1"
ttp <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
#
tmp <-data.frame(trait.pcaSq)
rownames(tmp) <- gsub("c1_", "M", tmp[,1])
tmp[,-1] <- apply(tmp[,-1],2, as.numeric)
#tmp <- tmp[o[1:num],]
tmp <- tmp[m,]
colnames(tmp)[1] <- "V1"
ttq <- reshape(tmp, idvar="V1", varying = list(names(tmp)[-1]), v.names="corr", timevar="Trait", times=names(tmp)[-1], direction = "long")
tt <- cbind(ttc, pval=ttp[,3], qval=ttq[,3])

tt <- tt[-grep("Corrected",tt$Trait),]             # remove "CorrectedSeroconversion"
tt[tt$Trait=="Composite_SeroConversion","Trait"] <- "Seroconversion"
tt[tt$Trait=="Composite_Baseline","Trait"] <- "Baseline HAI"
tt[tt$Trait=="Composite_D28","Trait"] <- "Response HAI"
tt$V1 <- gsub("c1_", "M", tt$V1)
```

### Preprocessed traits as alternative source without re-calculation

```{r}
#tt1: rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_trait_long.tsv")
#tt2: rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_avLog2Sero_long.tsv")

tt <- rbind (tl1, tl2)
```


### filter desired traits and modules

```{r}
num <- 10  # # to display
# Careful! some out columns are not numeric!
o <- order(out$avSero.rnk); ptt.t <- ""
m <- out[o[1:num],"module"]

ix <- which(!is.na(match(tt$V1, m)))
tt <- tt[ix,]

tt <- tt[-grep("omposite",tt$Trait),]             # remove all composite traits
tt[tt$Trait=="average_log2SC","Trait"] <- "Seroconversion"
tt[tt$Trait=="average_log2Baseline","Trait"] <- "Baseline HAI"
tt[tt$Trait=="average_log2D28","Trait"] <- "Response HAI"

```


### Draw combined DEG enrichment and correlation plot

```{r,fig.height=5,fig.width=8}
ptt <- ggplot(tt, aes(x=factor(Trait, levels=c("Age", "BMI", "Baseline HAI", "Response HAI", "Seroconversion")), y=factor(V1, levels=rev(m)))) +
  geom_point(aes(size=-log10(pval), color=corr)) + 
  scale_color_distiller(palette="RdYlBu", direction = -1, limits=c(-.5,.5)) + theme_bw() + 
  labs(x="Traits", y="Modules", title=ptt.t, size="-log10(P)") +
  theme(text = element_text(size = 14), title= element_text(size = 12), axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))

####
# use module enrich file
#tmp <- out[o[1:num],]
rownames(out) <- out$module
#tmp <- out[o[1:num],c(3,5:15)]
#tmp <- out[m,c(3,5:15)]  # old table
tmp <- out[m, c(4,6:16,54:56)]   # all trait DEG enrichments

tl <- reshape(tmp, idvar="module", varying = list(names(tmp)[-1]), v.names="pval", timevar="Trait", times=names(tmp)[-1], direction = "long")
tl <- tl[-grep("omposite",tl$Trait),]             # remove composite type traits
tl <- tl[tl$Trait!="Run.ID",]                      # remove "Run.ID"
tl <- tl[tl$Trait!="Race_Ethnicity",]               
tl <- tl[tl$Trait!="Month.vaccinated",]              
tl[tl$Trait=="av..Seroconversion","Trait"] <- "Seroconversion"
tl[tl$Trait=="av..Comp..Baseline","Trait"] <- "Baseline HAI"
tl[tl$Trait=="av..Comp..Response","Trait"] <- "Response HAI"

ptl <- ggplot(tl, aes(x=factor(Trait, levels=c("Age", "BMI", "Gender", "Prevacc.naive", "Baseline HAI", "Response HAI", "Seroconversion")), y=factor(module, levels=rev(m)), fill=-log10(as.numeric(pval)))) + 
  geom_tile() + scale_fill_distiller(palette="YlOrRd", direction = 1) + 
  labs(x="Traits", y="Modules", title="", fill="-log10(P)") +
  theme(text = element_text(size = 14), title= element_text(size = 12), axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))

ePlot <- list(ptl, ptt)

plot_grid(plotlist=ePlot, ncol=2, nrow=1)
```


# GSEA

## Download human annotations for GSEA

```{r}
require(DGCA)   # actually should get on top of markdown file.
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

## Create gene list ranked by Spearman correlation

```{r}
gs <- intersect(colnames(geno.xT), y$c1_198)
x.modT <- geno.xT[, gs]   # select the corresponding genes from module
corr <- cor(x.modT, trait$Composite_SeroConversion, method="spearman")

corr <- corr[order(-corr[,1]),]
gene.list <- corr
```



# Rund DGCA on select modules

```{r}
# do some c1_x <-> M* module name translation back and forth
m1 <- gsub("M", "c1_", m)
b.tbl <- List2Mod(b[m1])
b.tbl$Module <- gsub("c1_", "M", b.tbl$Module)

moduleGO_res <- moduleGO(genes=b.tbl$Values, labels=b.tbl$Module, universe=unique(genes[rnaseq.all.logCPM$X,"hgnc_symbol"]), pval_GO_cutoff = 0.05)

```


# Plot moduleGO results

```{r,fig.height=10,fig.width=10}
moduleGO_df <- extractModuleGO(moduleGO_res)
# shorten GO names
t2 <- strsplit(moduleGO_df$Term," ")
t3 <- unlist(lapply(t2,function(x) {p<-c(x[1:3],ifelse(length(x)>3,x[length(x)],""));
      ix<-which(is.na(p));if(length(ix)>0) {p<-p[1:(min(ix)-1)]};paste(p,collapse = " ")}))
moduleGO_df2 <- moduleGO_df
moduleGO_df2$Term <- t3 

# restrict P-value
moduleGO_df_le50 <- moduleGO_df2
moduleGO_df_le50[,grep("Val", colnames(moduleGO_df_le50))] <- apply(moduleGO_df_le50[,grep("Val", colnames(moduleGO_df_le50))], 2, function(x) ifelse(x<1e-50,1e-50,x))

moduleGO_df_le20 <- moduleGO_df2
moduleGO_df_le20[,grep("Val", colnames(moduleGO_df_le20))] <- apply(moduleGO_df_le20[,grep("Val", colnames(moduleGO_df_le20))], 2, function(x) ifelse(x<1e-20,1e-20,x))


pmGO <- plotModuleGO(moduleGO_df_le20, nTerms = 3, text_size = 10, coord_flip = T)

```



# Combine plots for Fig. 4

```{r,fig.height=12.5,fig.width=10.5}
p1 <- plot_grid(ptl, ptt, NULL, NULL, NULL, NULL, nrow =2, rel_heights = c(1.9,3), labels=c("A", "B", "C", "D", "", "E"))

ggdraw(p1) +
  draw_image("Fig4part3.png", x=.66, y=.34, width=0.31 ) +
  draw_image("MEGENA/M387a.png", x=.04, y=-0.1, width=0.545) +
  draw_plot(pmGO, x = .59, y=0.12, width=0.5, height = 0.4) +
  draw_image("color_scale.png", x=.1, y=-.3, width=.25) +
  draw_text("M387", x=.44, y=.31)
ggsave("Fig_4.png", height = 12.5, width = 10.5)
ggsave("Fig_4.pdf", height = 12.5, width = 10.5)
```


# Combine plots for Fig. S15 (Age dependence)

```{r,fig.height=12.5,fig.width=10.5}
p1 <- plot_grid(ptl, ptt, NULL, NULL, NULL, NULL, nrow =2, rel_heights = c(2.1,3), labels=c("A", "B", "D", "C", "", "E"))

ggdraw(p1) +
  draw_plot(pmGO, x=0.09, y=0.03, width=0.55, height=0.5) +
  draw_image("MEGENA/M91.png", x=.66, y=.3, width=0.35 ) +
  draw_image("MEGENA/M505.png", x = 1,
    hjust = 1, halign = 1, valign = .2, width=0.32 ) +
  draw_text("M91", x=.96, y=.8) +
  draw_text("M505", x=.7, y=.51)
ggsave("Fig_S15.png", height = 12.5, width = 10.5)
ggsave("Fig_S15.pdf", height = 12.5, width = 10.5)
```


# Combine plots for Fig. S16 (BMI dependence)

```{r,fig.height=12.5,fig.width=10.5}
p1 <- plot_grid(ptl, ptt, NULL, NULL, NULL, NULL, nrow =2, rel_heights = c(2.1,3), labels=c("A", "B", "D", "C", "", "E"))

ggdraw(p1) +
  draw_plot(pmGO, x=0.06, y=0.03, width=0.57, height=0.5) +
  draw_image("MEGENA/M47.png", x=.7, y=.3, width=0.3 ) +
  draw_image("MEGENA/M76.png", x = 1,
    hjust = 1, halign = 1, valign = .3, width=0.39 ) +
  draw_text("M47", x=.72, y=.8) +
  draw_text("M76", x=.91, y=.56) 
ggsave("Fig_S16.png", height = 12.5, width = 10.5)
ggsave("Fig_S16.pdf", height = 12.5, width = 10.5)
```




