---
title: "HAI Analysis Fig 6"
author: "Matthew Chung"
date: "12/19/2021"
output: 
  html_document:
    toc: true
    theme: united
---

# Preanalysis
## Load packages

```{r,warning=FALSE,message = FALSE}
library(biomaRt)
# library(cowplot)
# library(clusterProfiler)
# library(cowplot)
# library(DESeq2)
library(edgeR)
library(egg)
# library(enrichplot)
library(ggplot2)
# library(gprofiler2)
library(matrixStats)
# library(openxlsx)
# library(pathview)
library(pROC)
library(reshape2)
library(rlist)
# library(see)
library(splines)
# library(SuperExactTest)
# library(UpSetR)
```

## Load functions
```{r}
lm_eqn <- function(df){
    m <- lm(df[,2] ~ df[,1], df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));
}
```

## Load upstream data
```{r}
load("../data.RData")
```

## Set metadata covariates
```{r}
covar1 <- c("log2sero","seroconversion_2cat")
covar2 <- c("Age","Gender","BMI_Value","Prevacc.naive","Composite_Baseline")
## covar2 <- c("Age","Gender","BMI", "Race_Ethnicity","Month.vaccinated")

```

## Add metadata readout for categorizing Composite_Seroconversion
```{r}
metadata$log2sero <- log2(metadata$average_seroconversion)
metadata$seroconversion_2cat <- as.factor(ifelse(metadata$log2sero >= 2,"high","low"))
metadata$D0_score_2cat <- as.factor(ifelse(metadata$Composite_Baseline >= 25, "high", "low"))
```

# Create random forest with Composite_SeroConversion DE genes to predict Composite_SeroConversion category

## Create container for Fig 6 and 7

```{r}
figA <- list()
figB <- list()
```

## Run differential expression analysis

```{r}
FDR.cutoff <- 0.05

degenes.edgeR <- list()

for(j in 1:length(covar1)){
  group <- metadata[,which(colnames(metadata) == covar1[j])]
  if(length(unique(group)) > 10){
    spline <- ns(group, df=3)
    design <- model.matrix(~spline)
  }else{
    design <- model.matrix(formula(paste("~",covar1[j])),data = metadata)
  }
  
  y <- DGEList(counts=counts.subset,
               group=group)
  keep <- filterByExpr(y)  
  y <- y[keep, , keep.lib.sizes = F]
  y <- calcNormFactors(y, method="TMM")
  y <- estimateDisp(y,design, robust=TRUE)
  fit <- glmQLFit(y,design,robust=TRUE)
  
  qlf <- glmQLFTest(fit,coef=2:(ncol(design)))
  degenes <- topTags(qlf,n=Inf)
  degenes.edgeR[[j]] <- rownames(degenes$table)[degenes@.Data[[1]]$FDR < FDR.cutoff]
}
names(degenes.edgeR) <- covar1
```

## Switch TPM rownames to gene names and subset by Composite_SeroConversion DE genes

```{r}
ensembl <- useEnsembl(biomart = "genes")
ensembl <- useDataset(dataset = "hsapiens_gene_ensembl", mart = ensembl)
biomart <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name'),
                filters = 'ensembl_gene_id',
                values = degenes.edgeR$log2sero, 
                mart = ensembl)
biomart[,2] <- ifelse(biomart[,2] == "" | biomart[,2] == "Metazoa_SRP",biomart[,1],biomart[,2])
#tpm.subset <- tpm.subset[degenes.edgeR$Composite_SeroConversion,]
#rownames(tpm.subset) <- biomart[match(rownames(tpm.subset),biomart[,1]),2]
```
## Load modules

```{r}
modules <- list(PC1=read.delim("./data/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_modules_PC1.tsv"),
               PC2=read.delim("./data/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_modules_PC2.tsv"),
               PC3=read.delim("./data/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_modules_PC3.tsv"),
               PC4=read.delim("./data/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_modules_PC4.tsv"),     
               PC5=read.delim("./data/rnaseq-merged-counts_Matt_all_logCPM_RunIDMonthVacAdj_FDR0.05_modules_PC5.tsv"))

for(i in 1:length(modules)){
  rownames(modules[[i]]) <- modules[[i]][,2]
  modules[[i]] <- modules[[i]][,3:ncol(modules[[i]])]
  modules[[i]] <- modules[[i]][,metadata$File.Name]
}
```


## Create random forest model

### Split data into training and testing sets

```{r}
seeds <- c(80:89)
rf.alldata <- list(PD=as.data.frame(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2]),
                   PD_DEGs=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                t(tpm.subset[degenes.edgeR$log2sero,]))),
                   PD_dis=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                   t(tpm.subset[degenes.edgeR$seroconversion_2cat,]))),
                   PD_DEGs_PC1=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[degenes.edgeR$log2sero,]),
                                                    t(modules$PC1))),
                   PD_DEGs_PC2=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[degenes.edgeR$log2sero,]),
                                                    t(modules$PC2))),
                   PD_DEGs_PC3=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[degenes.edgeR$log2sero,]),
                                                    t(modules$PC3))),
                   PD_DEGs_PC4=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[degenes.edgeR$log2sero,]),
                                                    t(modules$PC4))),
                   PD_DEGs_PC5=as.data.frame(cbind(metadata[match(colnames(tpm.subset),metadata$File.Name),covar2],
                                                    t(tpm.subset[degenes.edgeR$log2sero,]),
                                                    t(modules$PC5))))

rf.traindata <- list()
rf.testdata <- list()

library(caTools)
for(i in 1:length(rf.alldata)){
  rf.traindata[[i]] <- list()
  rf.testdata[[i]] <- list()
  for(j in 1:length(seeds)){
    set.seed(seeds[j])
    sample = sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    rf.traindata[[i]][[j]] = rf.alldata[[i]][sample == T,]
    rf.testdata[[i]][[j]]  = rf.alldata[[i]][sample == F,]
  }
}
names(rf.traindata) <- names(rf.alldata)
names(rf.testdata) <- names(rf.alldata)
```
    
### Tune random forest model for optimal number of variables sampled at each split

```{r,warning=FALSE,message = FALSE}

rf.mdl <- list()

library(randomForest)
for(i in 1:length(rf.alldata)){
  rf.mdl[[i]] <- list()
  for(j in 1:length(seeds)){
    set.seed(seeds[j])
    sample <- sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    set.seed(seeds[j])
    mtry <- tuneRF(x = rf.traindata[[i]][[j]],
                   y = metadata$seroconversion_2cat[sample == T],
                   mtryStart=round(ncol(rf.traindata[[i]][[j]])/3),
                   ntreeTry=500,
                   stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
    best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
  
    set.seed(seeds[j])
    sample = sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    set.seed(seeds[j])
    rf.mdl[[i]][[j]] <-randomForest(x = rf.traindata[[i]][[j]],
                                y = metadata$seroconversion_2cat[sample == T],
                                mtry=best.m,
                                ntree=500,
                                proximity =T,
                                importance = T)
    print(paste(i,j))
  }
}
```

### Plot average OOB error

```{r,fig.height=3,fig.width=6}
colorsx <- c("#C988EE","#A3F38D","#8DE5F3")

col1 <- c()
col2 <- c()

for(i in 1:length(rf.alldata)){
  col1 <- c(col1,rep(names(rf.alldata)[i],10))
  col2 <- c(col2,unlist(lapply(rf.mdl[[i]],function(x){x$err.rate[500,1]})))
}
oob.error <- as.data.frame(cbind(col1,col2))

plot.df <- oob.error

plot.df[,2] <- as.numeric(as.character(plot.df[,2]))
#plot.df[,1] <- factor(plot.df[,1],levels=c("Metadata","Composite_SeroConversion DE genes","Metadata + Composite_SeroConversion DE genes"))

levels <- c("PD","PD_DEGs","PD_DEGs_PC1","PD_DEGs_PC2",
            "PD_DEGs_PC3","PD_DEGs_PC4","PD_DEGs_PC5")
plot.df <- plot.df[plot.df$col1 %in% levels,]
plot.df$col1 <- factor(plot.df$col1,levels = rev(levels))

oob.plot <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[,2],
                     y=!!plot.df[,1],
                     fill=plot.df[,1]),
                 #fill=colorsx,
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[,2],
                  y=!!plot.df[,1]))+
    guides(fill="none")+
    labs(x="OOB estimate of error rate",y="", fill = "")+
    theme_bw()

figA$a <- oob.plot
```

### Plot boxplot of AUC values from ROC curves

```{r,fig.height=3,fig.width=10}

col1 <- c()
col2 <- c()

for(i in 1:length(rf.mdl)){
  for(j in 1:length(rf.mdl[[i]])){
    set.seed(seeds[j])
    sample <- sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    col1 <- c(col1,names(rf.alldata)[i])
    col2 <- c(col2,
              auc(metadata$seroconversion_2cat[sample == F],
                  predict(rf.mdl[[i]][[j]], newdata=rf.testdata[[i]][[j]],type="prob")[,1]))
  }
}
plot.df.auc <- as.data.frame(cbind(col1,col2))

levels <- c("PD","PD_DEGs","PD_DEGs_PC1","PD_DEGs_PC2",
            "PD_DEGs_PC3","PD_DEGs_PC4","PD_DEGs_PC5")
plot.df.auc <- plot.df.auc[plot.df.auc$col1 %in% levels,]
plot.df.auc$col1 <- factor(plot.df.auc$col1,levels = rev(levels))
auc.boxplot <- ggplot()+
    geom_boxplot(aes(x=!!as.numeric(as.character(plot.df.auc[,2])),
                     y=!!plot.df.auc[,1],
                     fill=plot.df.auc[,1]),
                 #fill=colors,
                 
                 outlier.shape = NA)+
    geom_point(aes(x=!!as.numeric(as.character(plot.df.auc[,2])),
                  y=!!plot.df.auc[,1]))+
  guides(fill="none")+
    labs(x="AUC",y="", fill = "")+
    theme_bw()


figA$b <- auc.boxplot



pdf("figs6-1.pdf",
    height=3,
    width=12)
plot_grid(figA$a,figA$b+theme(axis.text.y=element_blank()),
               ncol=2)
dev.off()



```

### Plot confusion matrices

```{r,fig.height=5,fig.width=15}
reduceApplyListOfArrays<- function(x){
     y<-apply(array(unlist(x), c(dim(x[[1]]), dim(x[[2]]), length(x))), 
              c(1,2), mean)
     colnames(y)<-colnames(x[[1]])
     rownames(y)<-rownames(x[[1]])
  return(y)
}

confusion_matrix_plots.list <- list()
for(i in 1:length(rf.mdl)){
  matrix.list <- lapply(rf.mdl[[i]],function(x){return(x$confusion)})
  
  plot.df <- melt(reduceApplyListOfArrays(matrix.list)[1:2,1:2])
  
  plot.df[,1] <- factor(plot.df[,1],levels=c("low","high"))
  plot.df[,2] <- factor(plot.df[,2],levels=c("low","high"))
  
  confusion_matrix_plots.list[[i]] <- ggplot(plot.df , aes(x = Var2, y = Var1, fill = value)) +
    geom_tile(color = "white",
              lwd = 1.5,
              linetype = 1) +
    geom_text(aes(label = paste0(value,"\n",round(value/sum(value)*100,2),"%")), color = "black", size = 4) +
    scale_fill_gradient2(low = "#075AFF",
                         mid = "#FFFFCC",
                         high = "#FF0000") +
    labs(x="predicted values",y="actual values",fill="")+
    coord_fixed()+
    theme_minimal()
}

egg::ggarrange(plots=confusion_matrix_plots.list,
          labels = c("A","B","C","D","E"),
          ncol=3)
figA$c <- confusion_matrix_plots.list
```

### Plot ROC curves

```{r,fig.height=5,fig.width=10}
roc.list <- list()
auc <- c()
for(i in 1:length(rf.mdl)){
  roc.df <- as.data.frame(matrix(nrow=0,
                                 ncol=2))
  
  for(j in 1:length(rf.mdl[[i]])){
    set.seed(seeds[j])
    sample <- sample.split(metadata$seroconversion_2cat, SplitRatio = .75)
    
    roc.df <- as.data.frame(rbind(roc.df,
                                  cbind(metadata$seroconversion_2cat[sample == F],
                                        predict(rf.mdl[[i]][[j]], newdata=rf.testdata[[i]][[j]],type="prob")[,1])))
    # ,emetadata$seroconversion_2cat[sample == F],
    #                      predict(rf.mdl[[i]][[j]], newdata=rf.testdata[[i]][[j]],type="prob")[,1],
    #                      smooth=T)
  }
  roc.list[[i]] <- roc(roc.df[,1],roc.df[,2],smooth=T)
  auc[i] <- round(auc(roc.df[,1],roc.df[,2]),4)
}

ci.list <- lapply(roc.list, ci.se, specificities = seq(0, 1, l = 25))
dat.ci.list <- lapply(ci.list, function(ciobj) 
  data.frame(x = as.numeric(rownames(ciobj)),
             lower = ciobj[, 1],
             upper = ciobj[, 3]))

p <- ggroc(roc.list) + 
  theme_bw() + 
  geom_abline(slope=1, intercept = 1, linetype = "dashed", alpha=0.7, color = "grey") + 
  coord_equal()+
  #scale_color_manual(values=colors,labels=c(paste0("Metadata | AUC: ",auc[1]),
                                            #paste0("Log2 Seroconversion DE Genes | AUC: ",auc[2]),
                                            #paste0("Metadata + Log2 Seroconversion DE Genes | AUC: ",auc[3])))+
  labs(color="RF Model")+
  geom_line(size=1)

for(i in 1:3) {
  p <- p + geom_ribbon(
    data = dat.ci.list[[i]],
    aes(x = x, ymin = lower, ymax = upper),
    fill = i + 1,
    alpha = 0.2,
    inherit.aes = F) 
  } 

p
```




### Plot random forest variable importance

```{r,fig.height=10,fig.width=11}
colors <- c("#C988EE","#A3F38D","#8DE5F3")
colors <- c("#FB61D7","#A58AFF","white","#00B6EB","#00C094","#53B400","#C49A00","#F8766D",
            "white","white","white","white","white")
names <- c()

acc_plots.list <- list()
for(i in 1:length(rf.mdl)){
  importance.df <- do.call(cbind,lapply(rf.mdl[[i]],function(x){return(x$importance)}))
  rownames(importance.df) <- gsub("D0_ALLSTRAINS_SEROSTATUS","Initial HAI",rownames(importance.df))
  rownames(importance.df) <- gsub("Gender","Sex",rownames(importance.df))
  rownames(importance.df) <- gsub("Race_Ethnicity","Race",rownames(importance.df))
  rownames(importance.df) <- gsub("Month.vaccinated","Month Vaccinated",rownames(importance.df))
  rownames(importance.df) <- gsub("Prevacc.naive","Previous Vaccination\nStatus",rownames(importance.df))
  
  acc.order <- rownames(importance.df)[order(-rowMeans(importance.df[,colnames(importance.df) == "MeanDecreaseAccuracy"]))]
  gini.order <- rownames(importance.df)[order(-rowMeans(importance.df[,colnames(importance.df) == "MeanDecreaseGini"]))]
  
  if(i!=1){
    acc.order <- acc.order[1:10]
    gini.order <- gini.order[1:10]
  }
  
  if(i!=3){
  terms <- unique(grep("ENSG",c(acc.order,gini.order),value = T))
  if(length(terms) > 0){
    for(j in 1:length(terms)){
      old <- terms[j]
      new <- biomart[biomart[,1] == old,2]
      if(is.na(new)){old <- new}
      acc.order <- gsub(old,new,acc.order)
      gini.order <- gsub(old,new,acc.order)
      rownames(importance.df) <- gsub(old,new,rownames(importance.df))
    }
  }
  }
    
  plot.df <- melt(importance.df)
  plot.df <- plot.df[plot.df[,1] %in% acc.order,]
  plot.df[,1] <- factor(plot.df[,1],levels=rev(acc.order))
  
  plot.acc <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",3],
                     y=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",1]),
                 fill=colors[i],
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",3],
                   y=!!plot.df[plot.df[,2] == "MeanDecreaseAccuracy",1]))+
    labs(title=names(rf.alldata)[i],x="MeanDecreaseAccuracy",y="")+
    theme_bw()
  
  plot.df <- melt(importance.df)
  plot.df <- plot.df[plot.df[,1] %in% gini.order,]
  plot.df[,1] <- factor(plot.df[,1],levels=rev(gini.order))

  plot.gini <- ggplot()+
    geom_boxplot(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseGini",3],
                     y=!!plot.df[plot.df[,2] == "MeanDecreaseGini",1]),
                 fill=colors[i],
                 outlier.shape = NA)+
    geom_point(aes(x=!!plot.df[plot.df[,2] == "MeanDecreaseGini",3],
                   y=!!plot.df[plot.df[,2] == "MeanDecreaseGini",1]))+
    labs(title=names(rf.alldata)[i],x="MeanDecreaseGini",y="")+
    theme_bw()
  
   names <- c(names,paste0(names(rf.alldata)[i],"_1"),paste0(names(rf.alldata)[i],"_2"))
   acc_plots.list <- list.append(acc_plots.list,
                                 plot.acc,
                                 plot.gini)
   
}
names(acc_plots.list) <- names

acc_plots.list
pdf("figs17.pdf",
    height=10,
    width=11)
egg::ggarrange(acc_plots.list$PD_1,acc_plots.list$PD_2,
               acc_plots.list$PD_DEGs_1,acc_plots.list$PD_DEGs_2,
               acc_plots.list$PD_DEGs_PC1_1,acc_plots.list$PD_DEGs_PC1_2,
               acc_plots.list$PD_DEGs_PC2_1,acc_plots.list$PD_DEGs_PC2_2,
               acc_plots.list$PD_DEGs_PC3_1,acc_plots.list$PD_DEGs_PC3_2,
               acc_plots.list$PD_DEGs_PC4_1,acc_plots.list$PD_DEGs_PC4_2,
               acc_plots.list$PD_DEGs_PC5_1,acc_plots.list$PD_DEGs_PC5_2,
          ncol=4)
dev.off()
figB <- acc_plots.list

png("figs17.png",
    height=10,
    width=11,
    units = 'in',
    res = 300)
egg::ggarrange(acc_plots.list$PD_1,acc_plots.list$PD_2,
               acc_plots.list$PD_DEGs_1,acc_plots.list$PD_DEGs_2,
               acc_plots.list$PD_DEGs_PC1_1,acc_plots.list$PD_DEGs_PC1_2,
               acc_plots.list$PD_DEGs_PC2_1,acc_plots.list$PD_DEGs_PC2_2,
               acc_plots.list$PD_DEGs_PC3_1,acc_plots.list$PD_DEGs_PC3_2,
               acc_plots.list$PD_DEGs_PC4_1,acc_plots.list$PD_DEGs_PC4_2,
               acc_plots.list$PD_DEGs_PC5_1,acc_plots.list$PD_DEGs_PC5_2,
          ncol=4)
dev.off()


```

### Run crossvalidation for random forest

```{r,fig.height=5,fig.width=10}
library(rfUtilities)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

cols <- gg_color_hue(8)
cols <- rev(cols)

figA$d <- list()
for(i in 1:length(rf.mdl)){
  set.seed(86)
  rf.cv <- rfcv(trainx = rf.alldata[[i]], 
                trainy = metadata$seroconversion_2cat, 
                mtryStart=round(ncol(rf.alldata[[i]])/3),
                cv.fold=100)
  plot <- ggplot(mapping=aes(x=!!rf.cv$n.var,
                                         y=!!rf.cv$error.cv))+
                            geom_point(size=2, color = cols[i])+
                            geom_line(size=1, color = cols[i])+
                            labs(title=names(rf.alldata)[i],x="n.var",y="error.cv")+
                            scale_x_continuous(trans="log10")+
                            theme_bw()
  figA$d[[i]] <- plot
  print(i)
}

 
 

names(figA$d) <- names(rf.alldata)
pdf("figs6-2.pdf",
    height=5,
    width=10)
egg::ggarrange(figA$d$PD,figA$d$PD_DEGs,figA$d$PD_DEGs_PC1,figA$d$PD_DEGs_PC2,figA$d$PD_DEGs_PC3,figA$d$PD_DEGs_PC4,
               figA$d$PD_DEGs_PC5,
               ncol=4)
dev.off()

figA$d
```

# Print figures in Fig 6
```{r,fig.height=8,fig.width=11}

ggarrange(egg::ggarrange(figA$a,figA$b+theme(axis.text.y = element_blank()),nrow=1),
          egg::ggarrange(figA$c[[1]],figA$c[[2]],figA$c[[3]],nrow=1),
          egg::ggarrange(figA$d[[1]],figA$d[[2]],figA$d[[3]],nrow=1),
          heights=c(1,2,2),
          nrow=3)
```

# Print figures in Fig 7

```{r,fig.height=8,fig.width=20}
egg::ggarrange(plots = figA$e,
          nrow=1)
```

```{r}
s16A <- plot_grid(figA$a,figA$b+theme(axis.text.y=element_blank()),
               ncol=2)
s16B <- egg::ggarrange(figA$d$PD,figA$d$PD_DEGs,figA$d$PD_DEGs_PC1,figA$d$PD_DEGs_PC2,figA$d$PD_DEGs_PC3,figA$d$PD_DEGs_PC4,
               figA$d$PD_DEGs_PC5,
               ncol=3)

s16 <- plot_grid(s16A, s16B, nrow = 2, labels = c("A", "B"))

png(filename = "figs16.png",
    height = 9,
    width = 6.8,
    units = 'in',
    res = 300)
s16
dev.off()

pdf(paste0("figs16.pdf"),
    height = 9,
    width = 6.8)
s16
dev.off()
```