---
title: "Differential Expression Analysis of Day 0 Samples - Covariate Analysis"
author: "Megan Hockman"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(biomaRt)
library(cowplot)
library(clusterProfiler)
# library(DESeq2)
library(edgeR)
library(enrichplot)
library(gameofthrones)
library(ggpubr)
library(GOSemSim)
library(ggplot2)
library(ggridges)
# library(gprofiler2)
# library(openxlsx)
# library(pathview)
library(reshape2)
# library(see)
library(splines)
# library(SuperExactTest)
library(UpSetR)
library(png)
library(ggpubr)
library(dplyr)
library(stringr) 
library(grid)
library(extrafont)
library(extrafontdb)
```

###Load the Data
```{r}
load("../data.RData")

metadata$log2sero <- as.numeric(log2(metadata$average_seroconversion))
metadata$log2baseline <- as.numeric(log2(metadata$Composite_Baseline))

```

Add a metadata column for age ranges
```{r}
sort(unique(metadata$Age))
metadata <- metadata %>% mutate(Agerange = ifelse(Age <= 30, "18-30",
                           ifelse(Age <=60, "30 - 60", "60-85")))

metadata$bmibivar<- ifelse(metadata$BMI_Value < 25, "Normal", "Overweight")

```


```{r}
covariates <- list(log2sero=c("Age","Gender","BMI_Value"),
               log2baseline=c("Age","Gender", "BMI_Value")) 
fig3 <- list()

```

```{r}
FDR.cutoff <- 0.05

degenes.edgeR <- list(log2sero=c(),
                      log2baseline=c())
degenes.edgeR.coeff <- list(log2sero=c(),
                            log2baseline=c())
toptags <- list(log2sero=c(),
                log2baseline=c())
    
for(i in 1:length(degenes.edgeR)){
  ref.var <- names(degenes.edgeR)[i]
  sero.group <- metadata[,which(colnames(metadata) == ref.var)]
  
  for(j in 1:length(covariates[[i]])){
    group <- metadata[,which(colnames(metadata) == covariates[[i]][[j]])]
    design <- model.matrix(~sero.group:group)
    
    y <- DGEList(counts = counts.subset,
                 group = group)
    keep <- filterByExpr(y)
    y <- y[keep, ,keep.lib.sizes = F]
    y <- calcNormFactors(y, method = "TMM")
    y <- estimateDisp(y, design, robust = TRUE)
    fit <- glmQLFit(y, design, robust = TRUE)
    qlf <- glmQLFTest(fit, coef = grep(":", colnames(design))[1]:(ncol(design)))
    
    degenes <- topTags(qlf, n = Inf)
    toptags[[i]][[j]] <- degenes
    
    degenes.edgeR[[i]][[j]] <- rownames(degenes$table)[degenes@.Data[[1]]$FDR < FDR.cutoff]
    degenes.edgeR.coeff[[i]][[j]] <- qlf$coefficients
    
    
  }
  names(degenes.edgeR[[i]]) <- covariates[[i]]
  names(degenes.edgeR.coeff[[i]]) <- covariates[[i]]
  names(toptags[[i]]) <- covariates[[i]]
}
```




```{r}
names(degenes.edgeR[[1]]) <- c("Age", "Sex", "BMI")
names(degenes.edgeR[[2]]) <- c("Age", "Sex", "BMI")

fig3$a <- upset(fromList(degenes.edgeR[[1]]), 
                order.by = "freq",
                nsets = length(covariates[[1]]),
                nintersects = 9,
                text.scale = 5,
                point.size = 6,
                line.size = 1,
                mainbar.y.label = "Intersection Size", 
                sets.x.label = "Differentially\nExpressed Genes",
                set_size.show = TRUE) 


fig3$a

pdf(paste0("fig3a.pdf"),
    height=15,
    width=30)
fig3$a
grid.text("Seroconversion", x = 0.65, y = 0.95,
              gp = gpar(fontsize = 40))
dev.off()

png(paste0("fig3a.png"),
    height=15,
    width=30,
    units = "in",
    res = 300)
fig3$a
grid.text("Seroconversion", x = 0.65, y = 0.95,
              gp = gpar(fontsize = 40))
dev.off()

fig3$b <- upset(fromList(degenes.edgeR[[2]]), 
                order.by = "freq",
                nsets = length(covariates[[2]]),
                nintersects = 9,
                text.scale = 5,
                point.size = 6,
                line.size = 1,
                mainbar.y.label = "Intersection Size", 
                sets.x.label = "Differentially\nExpressed Genes",
                set_size.show = TRUE
                ) 


fig3$b

pdf(paste0("fig3b.pdf"),
    height=15,
    width=30)
fig3$b
grid.text("Baseline HAI", x = 0.65, y = 0.95,
              gp = gpar(fontsize = 40))
dev.off()

png(paste0("fig3b.png"),
    height=15,
    width=30,
    units = "in",
    res = 300)
fig3$b
grid.text("Baseline HAI", x = 0.65, y = 0.95,
              gp = gpar(fontsize = 40))
dev.off()

```
Export TopTags Tables
```{r}
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

toptagsexp <- list(toptags[[2]][[3]], toptags[[1]][[3]])


for(j in 1:length(toptagsexp)){
  genenames <- as.data.frame(matrix(nrow = nrow(toptagsexp[[j]]$table),
                                 ncol = 2))
  geneids <- rownames(!!toptagsexp[[j]]$table)
  genenames <- as.data.frame(getBM(filters= "ensembl_gene_id", 
                  attributes = c("ensembl_gene_id", "hgnc_symbol"),
                  values=geneids,
                  mart= mart))
  toptagsexp[[j]]$table$hgnc <- genenames$hgnc_symbol[match(rownames(toptagsexp[[j]]$table), genenames$ensembl_gene_id)]
  print(j)
}

toptagsexp[[2]]

write.csv(toptagsexp[[1]]$table, file = "TopTags_baselineHAI_BMI.csv", row.names = TRUE)

write.csv(toptagsexp[[2]]$table, file = "TopTags_SeroConversion_BMI.csv", row.names = TRUE)

```

Plot differential expression of Baseline HAI:BMI
```{r}
genes <- toptags[[2]][[3]]$table
genes <- genes[genes$FDR < 0.05,]
genes <- rownames(genes)

log2tpm.genes <- log2(tpm.subset[genes,])
plot.df <- melt(as.matrix(log2tpm.genes))
plot.df <- plot.df[plot.df[,3] != -Inf,]

x <- dplyr::count(plot.df, vars = Var1)
x <- filter(x, n > 50)
genesig <- as.character(x$vars)
log2tpm.genesig <- log2(tpm.subset[genesig,])
plot.df <- melt(as.matrix(log2tpm.genesig))


plot.df[,4] <- metadata$log2baseline[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$BMI_Value[match(plot.df[,2], metadata$File.Name)]
plot.df[,6] <- metadata$bmibivar[match(plot.df[,2], metadata$File.Name)]

mart <- useEnsembl(biomart = "ensembl", "hsapiens_gene_ensembl", mirror = "useast")

plot.list <- list()

for(j in 1:10){
  plot.df.subset <- plot.df[plot.df[,1] == genesig[j],]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  if(is.na(plot.df.subset$Var7)){plot.df.subset$Var5 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,6],
                                                      group = !!plot.df.subset[,6])) +
    geom_point() +
    scale_color_discrete(name = "BMI Category") +
    geom_smooth(method = "lm") +
    guides(fill = 'none') +
    labs(x = "Baseline HAI",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,7]) +
    theme_bw()+
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 10),
          aspect.ratio = 1)
  
}

plot.list

png(paste0("figs11.png"),
    height = 3000,
    width = 2000,
    units = 'px',
    res = 300)
plot_grid(plotlist = plot.list, ncol =2, align = 'hv')
dev.off()

```
Plot differential expression of Seroconversion:BMI
```{r}
genes <- toptags[[1]][[3]]$table
genes <- genes[genes$FDR < 0.05,]
genes <- rownames(genes)

log2tpm.genes <- log2(tpm.subset[genes,])

plot.df <- melt(as.matrix(log2tpm.genes))
plot.df <- plot.df[plot.df[,3] != -Inf,]
x <- dplyr::count(plot.df, vars = Var1)
x <- filter(x, n > 50)

genesig <- as.character(x$vars)
log2tpm.genesig <- log2(tpm.subset[genesig,])
plot.df <- melt(as.matrix(log2tpm.genesig))


plot.df[,4] <- metadata$log2sero[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$BMI_Value[match(plot.df[,2], metadata$File.Name)]
plot.df[,6] <- metadata$bmibivar[match(plot.df[,2], metadata$File.Name)]

mart <- useEnsembl(biomart = "ensembl", "hsapiens_gene_ensembl", mirror = "useast")

plot.list <- list()

for(j in 1:10){
  plot.df.subset <- plot.df[plot.df[,1] == genesig[j],]
  plot.df.subset <- plot.df.subset[plot.df.subset[,3] != -Inf,]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  if(is.na(plot.df.subset$Var7)){plot.df.subset$Var7 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,6],
                                                      group = !!plot.df.subset[,6])) +
    scale_color_discrete(name = "BMI Category") +
    geom_point() +
    geom_smooth(method = "lm") +
    guides(fill = 'none') +
    labs(x = "Seroconversion",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,7]) +
    theme_bw()+
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 10),
          aspect.ratio = 1)
  
}

plot.list

png(paste0("figs12.png"),
    height = 3000,
    width = 2000,
    units = 'px',
    res = 300)
plot_grid(plotlist = plot.list, ncol =2, align = 'hv')
dev.off()
```

Plot representative DE genes with respect to Baseline HAI and BMI
###Gene 1: KRT79
###Gene 2: IGLV4-60

```{r}
names(degenes.edgeR[[2]]) <- covariates[[2]]

genes <- c("ENSG00000185640","ENSG00000211639")

log2tpm.genes <- log2(tpm.subset[genes,])

plot.df <- melt(as.matrix(log2tpm.genes))

plot.df[,4] <- metadata$log2baseline[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$BMI_Value[match(plot.df[,2], metadata$File.Name)]
plot.df[,6] <- metadata$bmibivar[match(plot.df[,2], metadata$File.Name)]

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

plot.list <- list()
for(j in 1:length(genes)){
  plot.df.subset <- plot.df[plot.df[,1] == genes[j],]
  plot.df.subset <- plot.df.subset[plot.df.subset[,3] != -Inf,]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  if(is.na(plot.df.subset$Var7)){plot.df.subset$Var5 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,6],
                                                      group = !!plot.df.subset[,6])) +
    geom_point() +
    geom_smooth(method = "lm", aes(color = !!plot.df.subset[,6], fill = !!plot.df.subset[,6])) +
    guides(fill = 'none') +
    scale_color_viridis_d(name = "BMI Category")+
    scale_fill_viridis_d() +
    labs(x = "Baseline HAI",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,7]) +
    theme_bw()+
      theme(text = element_text(size = 20, family = "Arial"),
          aspect.ratio = 1)
  
}

fig3$c <- plot.list[[1]]
fig3$d <- plot.list[[2]]

pdf(paste0("fig3c.pdf"),
    height=5,
    width=8)
fig3$c
dev.off()

pdf(paste0("fig3d.pdf"),
    height=5,
    width=8)
fig3$d
dev.off()

png(paste0("fig3c.png"),
    height=1200,
    width=1700,
    units = "px",
    res = 300)
plot(fig3$c)
dev.off()

png(paste0("fig3d.png"),
    height=1200,
    width=1700,
    units = "px",
    res = 300)
plot(fig3$d)
dev.off()
```

Plot representative DE genes with respect to Seroconversion and BMI

```{r}
names(degenes.edgeR[[1]]) <- covariates[[1]]

genes <- c("ENSG00000146122","ENSG00000102109")

log2tpm.genes <- log2(tpm.subset[genes,])

plot.df <- melt(as.matrix(log2tpm.genes))

plot.df[,4] <- metadata$log2sero[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$BMI_Value[match(plot.df[,2], metadata$File.Name)]
plot.df[,6] <- metadata$bmibivar[match(plot.df[,2], metadata$File.Name)]

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

plot.list <- list()
for(j in 1:length(genes)){
  plot.df.subset <- plot.df[plot.df[,1] == genes[j],]
  plot.df.subset <- plot.df.subset[plot.df.subset[,3] != -Inf,]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  if(is.na(plot.df.subset$Var7)){plot.df.subset$Var5 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,6],
                                                      group = !!plot.df.subset[,6])) +
    geom_point() +
    geom_smooth(method = "lm", aes(color = !!plot.df.subset[,6], fill = !!plot.df.subset[,6])) +
    guides(fill = 'none') +
    scale_color_viridis_d(name = "BMI Category")+
    scale_fill_viridis_d() +
    labs(x = "Seroconversion",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,7]) +
    theme_bw()+
      theme(text = element_text(size = 20, family = "Arial"),
          aspect.ratio = 1)
  
}

plot.list

fig3$e <- plot.list[[1]]
fig3$f <- plot.list[[2]]

pdf(paste0("fig3e.pdf"),
    height=5,
    width=8)
fig3$e
dev.off()

pdf(paste0("fig3f.pdf"),
    height=5,
    width=8)
fig3$f
dev.off()

png(paste0("fig3e.png"),
    height=1200,
    width=1700,
    units = "px",
    res = 300)
plot(fig3$e)
dev.off()

png(paste0("fig3f.png"),
    height=1200,
    width=1700,
    units = "px",
    res = 300)
plot(fig3$f)
dev.off()
```

Perform GSEA Analysis : BMI
```{r,warning=FALSE,message = FALSE}
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

###This code generates a list of differentially expressed genes, ranked by spearman correlation coefficient, with respect to baseline HAI and BMI

```{r}
target.genes <- genes <- c("ENSG00000185640","ENSG00000211639")
gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts = gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$log2baseline[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$bmibivar[match(plot.df[,2], metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 3))

corr.df.overweight <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                           ncol = 3))

corr.df.normal <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 3))

corr.df.overweight <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                           ncol = 3))

corr.df[,1] <- unique(plot.df[,1])

x <- cor.test(plot.df.overweight[,4], plot.df.overweight[,3], method = "spearman", exact= FALSE)

x
for(i in 1:nrow(corr.df)) {
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  plot.df.normal <- plot.df.subset[plot.df.subset[,5] == "Normal",]
  plot.df.overweight <- plot.df.subset[plot.df.subset[,5] == "Overweight",]
  corr.df.normal[i,2] <- cor(plot.df.normal[,4], plot.df.normal[,3],method = "spearman")
  corr.df.overweight[i,2] <- cor(plot.df.overweight[,4], plot.df.overweight[,3], method = "spearman")
}

```


```{r}
target.genes <- degenes.edgeR[[2]][[3]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$log2baseline[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```

This code generates a list of differentially expressed genes, ranked by FC, with respect to baseline HAI and BMI
```{r}
target.genes <- toptags[[2]][[3]]$table
target.genes <- target.genes[target.genes$FDR < 0.05,]



gene.list <- target.genes$FDR
names(gene.list) <- rownames(target.genes)
gene.list <-  sort(gene.list, decreasing = TRUE)
```

```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
  gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05,
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 1)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_HAI <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_HAI$Description <- factor(gsea.diff.expression_HAI$Description, levels = gsea.diff.expression_HAI$Description[order(gsea.diff.expression_HAI$enrichmentScore)])

fig3$g <- gsea.diff.expression_HAI %>%
    #filter(p.adjust < 0.01) %>%
    #filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
    scale_fill_viridis_c(name = "P Value") +
    labs(title = "Baseline HAI and BMI", x = "Enrichment Score") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45, family = "Arial"),
          legend.text = element_text(size = 45),
          legend.key.height = unit(1, 'cm'),
          axis.text.y = element_text(size = 45))

fig3$g

pdf(paste0("fig3g.pdf"),
    height=30,
    width=35)
plot(fig3$g)
dev.off()

png(paste0("fig3g.png"),
    height=4000,
    width=10000,
    units = "px",
    res = 300)
plot(fig3$g)
dev.off()
```


###This code generates a list of differentially expressed genes with respect to Seroconversion/BMI
```{r}
target.genes <- degenes.edgeR[[1]][[3]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$log2sero[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```


This code generates a list of differentially expressed genes, ranked by FC, with respect to log2seroconversion and BMI
```{r}
target.genes <- toptags[[1]][[3]]$table
target.genes <- target.genes[target.genes$FDR < 0.05,]



gene.list <- target.genes$FDR
names(gene.list) <- rownames(target.genes)
gene.list <-  sort(gene.list, decreasing = TRUE)
```

Run GSEA of Seroconversion/BMI
```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05,
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 0.4)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_sero <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_sero$Description <- factor(gsea.diff.expression_sero$Description, levels = gsea.diff.expression_sero$Description[order(gsea.diff.expression_sero$enrichmentScore)])


fig3$h <- gsea.diff.expression_sero %>%
    #filter(p.adjust < 0.001) %>%
  filter(Description != "adaptive immune response based on somatic recombination of immune receptors built from immunoglobulin superfamily domains") %>%
  filter(Description != "positive regulation of cysteine-type endopeptidase activity involved in apoptotic process") %>%
    #filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
  labs(title = "Seroconversion and BMI", x = "Enrichment Score") +
    scale_fill_viridis_c(name = "P Value") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45, family = "Arial"),
          legend.text = element_text(size = 45),
          legend.key.height = unit(1, 'cm'),
          axis.text.y = element_text(size = 45))

fig3$h

pdf(paste0("fig3h.pdf"),
    height=30,
    width=35)
plot(fig3$h)
dev.off()

png(paste0("fig3h.png"),
    height=7000,
    width=12000,
    units = "px",
    res = 300)
plot(fig3$h)
dev.off()

fig3$h
```

Generate paper figures
```{r}
fig3DiffExp <- ((fig3$c | fig3$d) / (fig3$e | fig3$f))

pdf(paste0("fig3diffExp.pdf"),
    height=20,
    width=52)
plot(fig3GSEA)
dev.off()

png(paste0("fig3diffExp.png"),
    height=20,
    width=52,
    units = "in",
    res = 300)
plot(fig3GSEA)
dev.off()

fig3GSEA <- fig3$g + fig3$h

pdf(paste0("fig3GSEA.pdf"),
    height=20,
    width=52)
plot(fig3GSEA)
dev.off()

png(paste0("fig3GSEA.png"),
    height=20,
    width=52,
    units = "in",
    res = 300)
plot(fig3GSEA)
dev.off()


```

Plot all GSEA results in one plot
```{r}
gsea.diff.expression_HAI$group <- "HAI"
gsea.diff.expression_sero$group <- "Seroconversion"

gsea.diff.expression_all <- as_tibble(rbind(gsea.diff.expression_HAI, gsea.diff.expression_sero))

fig3_all <- gsea.diff.expression_all %>%
  filter(Description != "positive regulation of cysteine-type endopeptidase activity involved in apoptotic process") %>%
    ggplot(aes(enrichmentScore, Description, color = p.adjust, size = setSize)) +
    geom_point() +
  facet_grid(group~., scales = "free", space = "free")+
    labs(x = "Enrichment Score",
         y = "Biological Pathway",
         title = "") +
    scale_colour_viridis_c(name = "P Value") +
  scale_size(name = "Gene Set Size") +
    theme(text = element_text(size = 20)) 


pdf(paste0("fig3g_all.pdf"),
    height=12,
    width=15)
plot(fig3_all)
dev.off()

png(paste0("fig3g_all.png"),
    height=12,
    width=15,
    units = "in",
    res = 300)
plot(fig3_all)
dev.off()

```


From here on out, code will be optional - it creates plots and analyses of both baseline HAI and Seroconversion with respect to age and gender, rather than BMI which is shown in the paper.

Perform GSEA: Using age as the covariate

```{r,warning=FALSE,message = FALSE}
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

###This code generates a list of differentially expressed genes, ranked by spearman correlation coefficient, with respect to baseline HAI and age
```{r}
target.genes <- degenes.edgeR[[2]][[1]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$Composite_Baseline[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```


```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
  gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 0.7)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_HAI_Age <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_HAI_Age$Description <- factor(gsea.diff.expression_HAI_Age$Description, levels = gsea.diff.expression_HAI_Age$Description[order(gsea.diff.expression_HAI_Age$enrichmentScore)])

fig3$i <- gsea.diff.expression_HAI_Age %>%
    filter(p.adjust < 0.003) %>%
    filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
    scale_fill_viridis_c(name = "P Value") +
    labs(title = "Baseline HAI and Age", x = "Enrichment Score") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45),
          legend.text = element_text(size = 30),
          legend.key.height = unit(1, 'cm'))

fig3$i

pdf(paste0("fig3i_HAIGSEA.pdf"),
    height=20,
    width=22)
plot(fig3$i)
dev.off()


```


###This code generates a list of differentially expressed genes with respect to Seroconversion/Age
```{r}
target.genes <- degenes.edgeR[[1]][[1]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$Composite_SeroConversion[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```

Run GSEA
```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
  gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 0.7)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_sero_age <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_sero_age$Description <- factor(gsea.diff.expression_sero_age$Description, levels = gsea.diff.expression_sero_age$Description[order(gsea.diff.expression_sero_age$enrichmentScore)])


fig3$j <- gsea.diff.expression_sero_age %>%
    filter(p.adjust < 0.0005) %>%
    filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
  labs(title = "Seroconversion and Age", x = "Enrichment Score") +
    scale_fill_viridis_c(name = "P Value") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45),
          legend.text = element_text(size = 30),
          legend.key.height = unit(1, 'cm'))

pdf(paste0("fig3j_SeroGSEA.pdf"),
    height=20,
    width=25)
plot(fig3$j)
dev.off()


fig3$j
```

Perform GSEA: Using age as the covariate

```{r,warning=FALSE,message = FALSE}
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

###This code generates a list of differentially expressed genes, ranked by spearman correlation coefficient, with respect to baseline HAI and Gender
```{r}
target.genes <- degenes.edgeR[[2]][[2]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$Composite_Baseline[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```


```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
  gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 0.7)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_HAI_sex <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_HAI_sex$Description <- factor(gsea.diff.expression_HAI_sex$Description, levels = gsea.diff.expression_HAI_sex$Description[order(gsea.diff.expression_HAI_sex$enrichmentScore)])

fig3$k <- gsea.diff.expression_HAI_sex %>%
    filter(p.adjust < 0.0001) %>%
    filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
    scale_fill_viridis_c(name = "P Value") +
    labs(title = "Baseline HAI and Sex", x = "Enrichment Score") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45),
          legend.text = element_text(size = 30),
          legend.key.height = unit(1, 'cm'))

fig3$k

pdf(paste0("fig3k_HAIGSEA.pdf"),
    height=20,
    width=22)
plot(fig3$k)
dev.off()


```


###This code generates a list of differentially expressed genes with respect to Seroconversion/Gender
```{r}
target.genes <- degenes.edgeR[[1]][[2]]

gsea.counts <- counts.subset[rownames(counts.subset) %in% target.genes,]
gsea.counts <- gsea.counts[,colSums(gsea.counts) != 0]

y <- DGEList(counts=gsea.counts)
log2tpm <- log2(tpm.subset)
log2tpm <- log2tpm[rownames(log2tpm) %in% target.genes,]

plot.df <- melt(as.matrix(log2tpm))
plot.df[,4] <- metadata$Composite_SeroConversion[match(plot.df[,2],metadata$File.Name)]

corr.df <- as.data.frame(matrix(nrow = length(unique(plot.df[,1])),
                                ncol = 2))
corr.df[,1] <- unique(plot.df[,1])

for(i in 1:nrow(corr.df)){
  plot.df.subset <- plot.df[plot.df[,1] == corr.df[i,1],]
  corr.df[i,2] <- cor(plot.df.subset[,4],plot.df.subset[,3],method="spearman")
}

corr.df <- corr.df[order(-corr.df[,2]),]
gene.list <- corr.df[,2]
names(gene.list) <- corr.df[,1]

print(length(gene.list))
```

Run GSEA
```{r}
ontlist <- c("BP", "CC", "MF")
simplist <- list()
resultlist <- list()
tib.list <- list()

for (i in 1:length(ontlist)){
  hsGO <- godata('org.Hs.eg.db', ont = ontlist[i])
  set.seed(86)
  gse <- gseGO(geneList=gene.list, 
             ont = ontlist[i], 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = get(organism), 
             pAdjustMethod = "none",
             seed=T)
  simp <- clusterProfiler::simplify(gse, 
                                    semData = hsGO,
                                    cutoff = 0.7)
  simplist[[i]] <- simp
  tib.list[[i]] <- as_tibble(simp@result)
}

gsea.diff.expression_sero_sex <- as_tibble(rbind(tib.list[[1]], tib.list[[2]], tib.list[[3]]))

gsea.diff.expression_sero_sex$Description <- factor(gsea.diff.expression_sero_sex$Description, levels = gsea.diff.expression_sero_sex$Description[order(gsea.diff.expression_sero_sex$enrichmentScore)])


fig3$j <- gsea.diff.expression_sero_sex %>%
    filter(p.adjust < 0.0005) %>%
    filter(setSize > 6) %>%
    ggplot(aes(enrichmentScore, Description, size = setSize)) +
    geom_point(aes(fill = p.adjust, size = setSize), colour="black", pch=21) +
  labs(title = "Seroconversion and Sex", x = "Enrichment Score") +
    scale_fill_viridis_c(name = "P Value") +
    scale_size(name = "Gene Set Size", range = c(5, 25)) +
    theme(text = element_text(size = 45),
          legend.text = element_text(size = 30),
          legend.key.height = unit(1, 'cm'))

pdf(paste0("fig3l_SeroGSEA.pdf"),
    height=20,
    width=25)
plot(fig3$l)
dev.off()


fig3$l
```











This code plots Baseline HAI and SeroConversion vs. other covariates, if desired.


Plot differential expression of Baseline HAI:Sex
```{r}
genes <- toptags[[2]][[2]]$table
genes <- genes[genes$FDR < 0.05,]
genes <- rownames(genes)


y <- DGEList(counts = counts.subset)

log2tpm.genes <- log2(tpm.subset[genes,])

plot.df <- melt(as.matrix(log2tpm.genes))
plot.df <- plot.df[plot.df[,3] != -Inf,]
x <- count(plot.df, vars = Var1)
x <- filter(x, n > 100)

genesig <- as.character(x$vars)
log2tpm.genesig <- log2(tpm.subset[genesig,])
plot.df <- melt(as.matrix(log2tpm.genesig))

plot.df <- melt(as.matrix(log2tpm.genes))
plot.df[,4] <- metadata$Composite_Baseline[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$Gender[match(plot.df[,2], metadata$File.Name)]
#plot.df[,6] <- metadata$Agerange[match(plot.df[,2], metadata$File.Name)]

mart <- useEnsembl(biomart = "ensembl", "hsapiens_gene_ensembl", mirror = "useast")

plot.list <- list()

for(j in 1:30){
  plot.df.subset <- plot.df[plot.df[,1] == genesig[j],]
  plot.df.subset <- plot.df.subset[plot.df.subset[,3] != -Inf,]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  #if(is.na(plot.df.subset$Var5)){plot.df.subset$Var5 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,5],
                                                      group = !!plot.df.subset[,5])) +
    geom_point() +
    geom_smooth(method = "lm") +
    guides(fill = 'none') +
    labs(x = "Baseline HAI",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,6]) +
    theme_bw()+
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 10),
          aspect.ratio = 1)
  
}

plot.list
```


Plot differential expression of Seroconversion:Age
```{r}
genes <- toptags[[1]][[1]]$table
genes <- genes[genes$FDR < 0.05,]
genes <- rownames(genes)


y <- DGEList(counts = counts.subset)

log2tpm.genes <- log2(tpm.subset[genes,])

plot.df <- melt(as.matrix(log2tpm.genes))
plot.df <- plot.df[plot.df[,3] != -Inf,]
x <- count(plot.df, vars = Var1)
x <- filter(x, n > 100)

genesig <- as.character(x$vars)
log2tpm.genesig <- log2(tpm.subset[genesig,])
plot.df <- melt(as.matrix(log2tpm.genesig))

plot.df[,4] <- metadata$log2sero[match(plot.df[,2], metadata$File.Name)]
plot.df[,5] <- metadata$Age[match(plot.df[,2], metadata$File.Name)]
plot.df[,6] <- metadata$Agerange[match(plot.df[,2], metadata$File.Name)]

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

plot.list <- list()

for(j in 1:30){
  plot.df.subset <- plot.df[plot.df[,1] == genesig[j],]
  plot.df.subset <- plot.df.subset[plot.df.subset[,3] != -Inf,]
  
  genelist <- getBM(filters= "ensembl_gene_id", 
                  attributes= c("ensembl_gene_id","hgnc_symbol"),
                  values=plot.df.subset$Var1,
                  mart= mart)
  
  plot.df.subset$Var7 <- genelist$hgnc_symbol[match(plot.df.subset$Var1, genelist$ensembl_gene_id)]
  #if(is.na(plot.df.subset$Var5)){plot.df.subset$Var5 <- plot.df.subset$Var1[1]}
  
  plot.list[[j]] <- ggplot(data = plot.df.subset, aes(x = !!plot.df.subset[,4],
                                                      y = !!plot.df.subset[,3],
                                                      color = !!plot.df.subset[,6],
                                                      group = !!plot.df.subset[,6])) +
    geom_point() +
    geom_smooth(method = "lm") +
    guides(fill = 'none') +
    labs(x = "Seroconversion",
         y = expression("log"[2]*TPM),
         title = plot.df.subset[,7]) +
    theme_bw()+
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 10),
          aspect.ratio = 1)
  
}

plot.list
```
